---
title: "Objective_2"
author: "Diva Medina Camp"
date: "2023-05-21"
output: html_document
---

### Objective 2 - Narrowing Down Hot Spots

Seeing the global map of COVID-19 cases results in the stark realization that some countries are more affected than others. In order to narrow down your studies, create a table using kable from knitr listing the top countries as far as confirmations and deaths (sum values for provinces of the same country into one value and show the country only). 

```{r libraries, include=FALSE}

library(dplyr)
library(knitr)
library(kableExtra)
library(tidyverse)
library(tinytex)
```

```{r data_prep}

#extract data of global confirmed COVID cases
confirmed_COVID_global <- 
  read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv", stringsAsFactors=FALSE)

#extract data of global COVID deaths
deaths_COVID_global <- 
  read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv", stringsAsFactors=FALSE)

#variable holding dataframe with necessary columns for confirmed
lastday_confirmed <- 
  select(confirmed_COVID_global, Province.State, Country.Region, 
         "Latest_Day"=tail(names(confirmed_COVID_global),1))

lastday_confirmed <- lastday_confirmed[-c(107 ,245 ,286),]

#variable holding dataframe with necessary columns for deaths
lastday_deaths <-
  select(deaths_COVID_global, Province.State, Country.Region, 
         "Latest_Day"=tail(names(deaths_COVID_global),1))

lastday_deaths <- lastday_deaths[-c(107 ,245 ,286),]

COVID_confirmed <- lastday_confirmed %>%
  group_by(Country.Region) %>%
  summarize(sum(Latest_Day)) %>%
  rename("Counts"="sum(Latest_Day)", "Country"="Country.Region") %>%
  arrange(desc(Counts))

COVID_confirmed$Counts <- prettyNum(COVID_confirmed$Counts, big.mark = ',', scientific=FALSE)

COVID_deaths <- lastday_deaths %>%
  group_by(Country.Region) %>%
  summarize(sum(Latest_Day)) %>%
  rename("Counts"="sum(Latest_Day)", "Country"="Country.Region") %>%
  arrange(desc(Counts))

COVID_deaths$Counts <- prettyNum(COVID_deaths$Counts, big.mark = ',', scientific=FALSE)

COVID_confirmed <- COVID_confirmed %>% mutate(Rank = row_number()) 

col_order <- c(3,1,2)
COVID_confirmed <- COVID_confirmed[, col_order]

```

```{r table}
COVID_Table <- cbind(COVID_confirmed, COVID_deaths)

COVID_Table %>%
  kbl(align=c(rep("r",1), rep("l",1), rep("r",1), rep("l",1), rep("r",1)), 
      caption = "Table of Top Countries") %>%
  kable_styling(bootstrap_options = "striped", "responsive",
                full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Confirmations" = 2, "Deaths" = 2)) %>%
  scroll_box(width = "800px", height = "450px")

```


##### Diva Medina Camp
###### 05/21/2023









