---
title: "Project 2"
author: "Diva Medina Camp, Wesley Newcomb."
date: "2023-05-20<br><br>"
output: html_document
---

<style type="text/css">

h1.title {
  font-size: 38px;
  text-align: center;
}
h4.author { /* Header 4 - and the author and data headers use this too  */
    font-size: 18px;
  text-align: center;
}
h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  text-align: center;
}
<br>
</style>

```{r libraries, include=FALSE}

library(dplyr)
library(leaflet)
library(knitr)
library(kableExtra)
library(tinytex)
library(magrittr)
library(readr)
library(geosphere)
library(RCurl)
library(ggplot2)
library(tidyr)
library(stringr)
library(cowplot)
library(scales)
library(numform)
library(leaflet)
```

### Background
> The World Health Organization has recently employed a new data science initiative, *CSIT-165*, that uses data science to characterize pandemic diseases. 
*CSIT-165* disseminates data driven analyses to global decision makers.

> *CSIT-165* is a conglomerate comprised of two fabricated entities: *Global Health Union (GHU)* and *Private Diagnostic Laboratories (PDL)*. 
Your and your partner's role is to play a data scientist from one of these two entities.

### Data
> [2019 Novel Coronavirus COVID-19 (2019-nCoV) Data Repository by John Hopkins CSSE](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series)

> Data for 2019 Novel Coronavirus is operated by the John Hopkins University Center for Systems Science and Engineering (JHU CSSE).
Data includes daily time series CSV summary tables, including confirmations, recoveries, and deaths. 
Country/region are countries/regions hat conform to World Health Organization (WHO).
Lat and Long refer to coordinates references for the user. 
Date fields are stored in MM/DD/YYYY format.

### Instructions
> The Global Health Initiative has recently employed a new data science response team, CSIT-165, that uses data science to characterize pandemic diseases. CSIT-165 disseminates data driven analyses to global and local decision makers.

> CSIT-165 is a conglomerate comprised of two fabricated entities: World Health Organization (WHO) and U.S. Pandemic Response Team (USPRT). Your and your partner’s role is to play a data scientist from one of these two entities. Discuss with your partner to decide who will be part of WHO and USPRT.

> Getting Started
One project member per group must create a new repository on GitHub. Initialize this repository with a readme.md file that lists each member of the group. If your group decides to collaborate using a centralized workflow (recommended), then the project member that created the repository must declare their partners as collaborators in GitHub. Each project member will clone this repository onto their machine using RStudio. In RStudio, create a project from version control with GitHub using the HTTP address of the repository created by project member.

> All project members must first contribute to analyses by uploading data sets respective to the entity they belong to in the CSIT-165 data science response team. If you belong to WHO then you are responsible for providing code necessary to download the global data sets and if you are assigned to USPRT then you are responsible for providing code necessary to download the US data sets.

# Objectives
> This project will encompass many of the lessons we have learned throughout the course, including interactive visualizations. RMarkdown files must be written such that each time you render the document it will download the necessary data sets for analysis. Please render the RMarkdown file the day it is due to reflect the most recent data sets. With this added functionality, your code must be able to analyze the datasets regardless of the date you render your document. Unlike others projects in the past, you will have the ability to solve these problems using any method you choose. Be careful, however, with the methods you use as you will be graded on the appropriateness of your solution and how well you execute your desired algorithm. If there appears to be a logic fail in how you executed your code, you will be penalized. This is an opportunity to really showcase your new found skills in data science with R!

## Project Objectives

### Objective 1 - Global Map

You are tasked to create a world map to gain an appreciation for where the most occurrences of COVID-19 confirmations and deaths are located.

Create this map using leaflet for the most recent date as shown below. For this map, sum the confirmations and deaths of provinces into one value to depict the total number for the country they belong to. When creating a marker for each country in the map, calculate lat and long as the mean values for the provinces that make up each country.

Customize the map to reflect the differences in magnitude for confirmations and deaths. In the example map below, circle markers that are blue represent low values, gray represents neutral values, and red represents high values. Low, middle, and high values were categorized to aesthetically map the markers based on their probabilistic distribution using the quartile function. You may use any method you like so that it is logical and allows visualization of value intensity. As well, customize the map to include hover labels that indicate country names and popup labels to show the value of confirmations and deaths for that country.

```{r data prep}
#Author: Diva M Camp

#extract data of global confirmed COVID cases
confirmed_COVID_global <- 
  read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv", stringsAsFactors=FALSE)

#extract data of global COVID deaths
deaths_COVID_global <- 
  read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv", stringsAsFactors=FALSE)

#variable holding dataframe with necessary columns for confirmed
lastday_confirmed <- 
  select(confirmed_COVID_global, Province.State, Country.Region, Lat, Long, "Confirmations"=tail(names(confirmed_COVID_global),1))

#variable holding dataframe with necessary columns for deaths
lastday_deaths <-
  select(deaths_COVID_global, Province.State, Country.Region, Lat, Long, "Deaths"=tail(names(deaths_COVID_global),1))

#remove unnecesary data points like Olympics and cruise ship.
lastday_confirmed <- lastday_confirmed[-c(107 ,245 ,286),]
lastday_deaths <- lastday_deaths[-c(107 ,245 ,286),]

```

```{r obj_1}
#Create dataframe containing sum of deaths and confirmations by province
COVID_dataSum <- lastday_confirmed %>% 
  inner_join(lastday_deaths, by=c("Province.State", "Country.Region", "Lat", "Long")) %>%
  mutate(All = rowSums(select(., "Confirmations", "Deaths")))

#Create dataframe which averages lats and longs, and adds confirmations and deaths.
COVID_dataSum_country <- 
  group_by(COVID_dataSum, Country.Region) %>%
  summarize(Lat = mean(Lat,na.rm=TRUE), 
            Long = mean(Long,na.rm=TRUE), 
            Confirmations = sum(Confirmations), 
            Deaths = sum(Deaths), 
            All = sum(All))
  
#format numeric data to have commas
COVID_dataSum_country$Confirmations <- prettyNum(COVID_dataSum_country$Confirmations, big.mark=",", scientific = FALSE)
COVID_dataSum_country$Deaths <- prettyNum(COVID_dataSum_country$Deaths, big.mark=",", scientific = FALSE)


#color palette function
pal = colorQuantile(palette=c("dodgerblue3", "grey35", "red"), domain=COVID_dataSum_country$All, n=5)

#Generate Map 
leaflet(COVID_dataSum_country) %>%
  addTiles() %>%
  setView(lng=0, lat=0, zoom=1) %>%
  addCircleMarkers(lng=~Long, lat=~Lat, radius=4, 
                   label=~Country.Region, color=~pal(COVID_dataSum_country$All), 
                   group="Confirmations",
                   popup=paste("Confirmed: ", as.character(COVID_dataSum_country$Confirmations))) %>%
  addCircleMarkers(lng=~Long, lat=~Lat, radius=4, 
                   label=~Country.Region, color=~pal(COVID_dataSum_country$All), 
                   group="Deaths",
                   popup=paste("Deaths: ", as.character(COVID_dataSum_country$Deaths))) %>%
  addLayersControl(overlayGroups = c("Confirmations", "Deaths"),
                   options = layersControlOptions(collapsed = FALSE))

```


### Objective 2 - Narrowing Down Hot Spots

Seeing the global map of COVID-19 cases results in the stark realization that some countries are more affected than others. In order to narrow down your studies, create a table using kable from knitr listing the top countries as far as confirmations and deaths (sum values for provinces of the same country into one value and show the country only). 

```{r obj_2}

#Make dataframe for confirmations data. Group by country, sum, add commas to numbers.
COVID_confirmed <- lastday_confirmed %>%
  group_by(Country.Region) %>%
  summarize(sum(Confirmations)) %>%
  rename("Counts"="sum(Confirmations)", "Country"="Country.Region") %>%
  arrange(desc(Counts))

COVID_confirmed$Counts <- prettyNum(COVID_confirmed$Counts, big.mark = ',', scientific=FALSE)

#Make dataframe for deaths data. Group by country, sum, arrange in desc order, add commas to numbers.
COVID_deaths <- lastday_deaths %>%
  group_by(Country.Region) %>%
  summarize(sum(Deaths)) %>%
  rename("Counts"="sum(Deaths)", "Country"="Country.Region") %>%
  arrange(desc(Counts))

COVID_deaths$Counts <- prettyNum(COVID_deaths$Counts, big.mark = ',', scientific=FALSE)

#Add rank numbers to table, and order columns.
COVID_confirmed <- COVID_confirmed %>% mutate(Rank = row_number()) 

col_order <- c(3,1,2)
COVID_confirmed <- COVID_confirmed[, col_order]

#combine confirmations and deaths tables.
COVID_Table <- cbind(COVID_confirmed, COVID_deaths)

#Format table
COVID_Table %>%
  kbl(align=c(rep("r",1), rep("l",1), rep("r",1), rep("l",1), rep("r",1)), 
      caption = "Table of Top Countries") %>%
  kable_styling(bootstrap_options = "striped", "responsive",
                full_width = FALSE) %>%
  add_header_above(c(" " = 1, "Confirmations" = 2, "Deaths" = 2), font_size="large") %>%
  scroll_box(width = "800px", height = "550px")

```

##### CSIT-165, Spring 2023.